### 2017-07-21 manuscript.rmd
incorporated Stefan's comments

Over the last three decades, a rich set of phylogenetic comparative methods has been developed to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; @Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew; @Jhwueng:2013fi]. A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share more traits because they have a more recent common ancestor. This violates the fundamental assumption of observation independence in standard statistical methods. Phylogenetic comparative methods address this dependence. They have largely been applied to morphological and ecological traits, but are just as relevant to functional genomics [@Barker:2005be]. Even so, most comparative functional genomic studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons across species (Figure 1*A*). This leaves comparative functional genomic studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

@@

KMRR [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture. The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event). The ortholog conjecture has important biological and technical implications. It shapes our understanding of the functional diversity of gene families. It is also used to relate findings from well-studied genes to homologous genes that have not been investigated in detail. It has be applied to many trait of genes, from gene sequence to biochemical properties to expression. While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.

Despite its importance, there have been relatively few tests of the ortholog conjecture. Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz]. Analyses of domain structure were consistent with the ortholog conjecture [@Forslund:2011cw]. There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is the most thorough such expression study to date.

@@

We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] dataset onto `r n_gene_trees` gene trees parsed from ENSEMBL Compara [@Herrero:2016jj]. These are the same pre-computed trees that the orthology/ paralogy annotations KMRR used are based on. `r length(gene_trees_pruned)` gene trees passed taxon sampling criteria (`r min_genes_with_expression` genes) after removing tips without Tau values and had at least one speciation event. Of these, `r length(gene_trees_calibrated)` were successfully time calibrated. These calibrated trees were used to calculate phylogenetic independent contrasts for `r nodes_contrast %>% filter(D=="Y") %>% nrow()` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow()` speciation nodes. One of these trees is presented in *SI Appendix* Figure 4 to demonstrate the analysis.

@@

An additional concern is that within-species expression variation is not considered when making comparisons of Tau across homologous genes. Neglecting within species variation is expected to mislead phylogenetic comparative analyses in some cases [@Ives:2007cw; Felsenstein:2008cl; Rohlfs:2015bd]. Since branch length describes expected trait variation [@Felsenstein:1985ua], within species variation can be modeled by extending the terminal branches of each gene phylogeny [@Ives:2007cw]. We modeled within-species variation by extending each terminal branch by 7 million years. This value was selected because it corresponds to the age of the clade Hominini (defined by the most recent common ancestor of humans and chimpanzees), the shallowest node in the phylogeny of examined species. This provides an opportunity to assess the impact that extreme within-species variation (*i.e.*, as large as the variance between humans and chimpanzees and their most recent common ancestor) would have on our findings. These analyses still do not reject the null hypothesis, and do not provide support for the ortholog conjecture (*SI Appendix*, Figure 6).

@@

Since greater rates of expression evolution following duplication do not explain the lower correlation of Tau values in the pairwise comparison plots (Figure 2 *A, C, E*), this pattern must reflect some other property that is shared across these analyses. We found that the lower correlations for duplication comparisons can largely be explained by the greater variance in the age of duplication nodes. Each point in the pairwise correlation plot (Figure 2 *A, C, E*) summarizes multiple pairwise comparisons across nodes of a given event type (speciation or duplication, as indicated by color) and clade in the species tree (Theria, Mammalia, Amniota, etc..., as indicated by the age of the clade along the x axis). These clade names in turn are derived from the node annotations in the ENSEMBL Compara trees, which apply clade names to both speciation and duplication nodes. While speciation nodes in the gene trees have a clear correspondence to clades in the species trees, duplication nodes do not since duplication can occur at any point along any branch in the species tree. Neighboring clade ages are therefore a very rough approximation of the age of duplication events. This is apparent when node ages derived from calibrated trees are plotted against the age of the clade annotations for each node (*SI Appendix*, Figure 7*A*). Because duplication events that are all annotated with the same clade name can occur at very different times, pairwise comparisons across these nodes capture evolutionary changes in Tau across very different different branch lengths. They therefore have lower correlation than pairwise comparisons made across speciation nodes. Of the `r n_dup_original` duplication nodes considered here, `r n_dup_remaining` have a node age (as determined by the time calibrated trees) within `r age_window * 100`% of their clade annotation. If only these duplication events closest to their annotated clade nodes are retained, the pattern of lower correlation of Tau evolution for duplication events disappears for the dataset simulated under the null model (*SI Appendix*, Figure 7*B*). This correction, though, comes at the high cost of discarding `r round( (n_dup_original - n_dup_remaining )/ n_dup_original * 100, 1 )`% of duplication nodes.

@@

Once we addressed the issues above with the implementation of pairwise comparisons, we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between species. We found that all five of the lowest values in the mid phase distribution (Figure 3*A*) are for pairwise comparisons that include the ctenophore (comb jelly). When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (p = 0.14 for the early-mid comparison and p < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Figure 3b). This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua; @Ackerly:1999iu]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch. The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by changes along some of the same branches (Figure 1*A*). This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree. The number of comparisons impacted by each change depends on the structure of the phylogenetic tree, *i.e.* how the species are related to each other. Phylogenetic comparative methods were developed specifically to address this problem [@Felsenstein:1985ua].

@@

After parsing the trees from the Compara file with treeio, which was recently split from ggtree [@Yu:2016fo], we added Tau estimates generated by the KMRR `Rscript.R` to the tree data objects. We then pruned away tips without expression data, retaining only the trees with `r min_genes_with_expression` or more tips. We also only retained trees with one or more speciation events, as speciation events are required for calibration steps. This removes trees that have multiple genes from only one species after pruning away tips without expression data.

@@

```{r within_species_variation, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.width=6.5, fig.height=3, fig.cap="Investigation of the potential impact of within-species varianace. These plots correspond to Figures 2*A* and 2*B*, but with increased expected within-species variation modeled by increasing the terminal gene phylogeny branch lengths. (*A*) Pairwise comparisons still recover higher correlation across speciation events than duplication events. (*B*) There is not an excess of larger contrasts for duplication events and the analysis does not reject the null hypothesis. Modeling a large amount of within-species variation does not lead to support for the ortholog conjecture."}

>>>>

Over the last three decades, a rich set of phylogenetic comparative methods has been developed to address the challenges and opportunities of trait comparisons across species [@Felsenstein:1985ua; @Grafen:1989um; @Pagel:1999gc; @Revell:2012jd; @FitzJohn:2012ey; @Uyeda:2014jl; @Garamszegi:2014ew; @Jhwueng:2013fi; @Rohlfs:2015bd]. A central challenge is the dependence of observations across species due to the evolutionary history of species -- more closely related species share more traits because they have a more recent common ancestor. This violates the fundamental assumption of observation independence in standard statistical methods. Phylogenetic comparative methods address this dependence. They have largely been applied to morphological and ecological traits, but are just as relevant to functional genomics [@Barker:2005be]. Even so, most comparative functional genomic studies have abstained from phylogenetic approaches and instead rely on multiple pairwise comparisons across species (Figure 1*A*). This leaves comparative functional genomic studies susceptible to statistical problems and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

@@

KMRR [@KryuchkovaMostacci:2016iw] sought to test the ortholog conjecture. The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event). The ortholog conjecture has important biological and technical implications. It shapes our understanding of the functional diversity of gene families. It is also used to relate findings from well-studied genes to homologous genes that have not been investigated in detail. It has been applied to many traits of genes, from gene sequence to biochemical properties to expression. While the ortholog conjecture describes a specific pattern of functional diversity across genes, it is also articulated as a hypothesis about the process of evolution-- that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.

Despite its importance, there have been relatively few tests of the ortholog conjecture. Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz]. Analyses of domain structure were consistent with the ortholog conjecture [@Forslund:2011cw]. There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz], and KMRR is one of the most detailed to date.

@@

We mapped the Tau values calculated by KMRR [@KryuchkovaMostacci:2016iw] for the Brawand *et al.* 2011 [@Brawand:2011du] dataset onto `r n_gene_trees %>% format(big.mark=",")` gene trees parsed from ENSEMBL Compara [@Herrero:2016jj]. These are the same pre-computed trees that the orthology/ paralogy annotations KMRR used are based on. `r length(gene_trees_pruned) %>% format(big.mark=",")` gene trees passed taxon sampling criteria (`r min_genes_with_expression %>% format(big.mark=",")` genes) after removing tips without Tau values and had at least one speciation event. Of these, `r length(gene_trees_calibrated)  %>% format(big.mark=",")` were successfully time calibrated. These calibrated trees were used to calculate phylogenetic independent contrasts for `r nodes_contrast %>% filter(D=="Y") %>% nrow() %>% format(big.mark=",")` duplication nodes and `r nodes_contrast %>% filter(D=="N") %>% nrow() %>% format(big.mark=",")` speciation nodes. One of these trees is presented in *SI Appendix* Figure 4 to demonstrate the analysis.

@@

An additional concern is that within-species expression variation is not considered when making comparisons of Tau across homologous genes. Neglecting within species variation is expected to mislead phylogenetic comparative analyses in some cases [@Ives:2007cw; @Felsenstein:2008cl; @Rohlfs:2015bd]. Since branch length describes expected trait variation [@Felsenstein:1985ua], within species variation can be modeled by extending the terminal branches of each gene phylogeny [@Ives:2007cw]. We modeled within-species variation by extending each terminal branch by 7 million years. This value was selected because it corresponds to the age of the clade Hominini (defined by the most recent common ancestor of humans and chimpanzees), the shallowest node in the phylogeny of examined species. This provides an opportunity to assess the impact that extreme within-species variation (*i.e.*, as large as the variance between humans and chimpanzees and their most recent common ancestor) would have on our findings. These analyses still do not reject the null hypothesis, and do not provide support for the ortholog conjecture (*SI Appendix*, Figure 6).

@@

Since greater rates of expression evolution following duplication do not explain the lower correlation of Tau values in the pairwise comparison plots (Figure 2 *A, C, E*), this pattern must reflect some other property that is shared across these analyses. We found that the lower correlations for duplication comparisons can largely be explained by the greater variance in the age of duplication nodes. Each point in the pairwise correlation plot (Figure 2 *A, C, E*) summarizes multiple pairwise comparisons across nodes of a given event type (speciation or duplication, as indicated by color) and clade in the species tree (Theria, Mammalia, Amniota, etc..., as indicated by the age of the clade along the x axis). These clade names in turn are derived from the node annotations in the ENSEMBL Compara trees, which apply clade names to both speciation and duplication nodes. While speciation nodes in the gene trees have a clear correspondence to clades in the species trees, duplication nodes do not since duplication can occur at any point along any branch in the species tree. Neighboring clade ages are therefore a very rough approximation of the age of duplication events. This is apparent when node ages derived from calibrated trees are plotted against the age of the clade annotations for each node (*SI Appendix*, Figure 7*A*). Because duplication events that are all annotated with the same clade name can occur at very different times, pairwise comparisons across these nodes capture evolutionary changes in Tau across very different different branch lengths. They therefore have lower correlation than pairwise comparisons made across speciation nodes. Of the `r n_dup_original %>% format(big.mark=",")` duplication nodes considered here, `r n_dup_remaining %>% format(big.mark=",")` have a node age (as determined by the time calibrated trees) within `r age_window * 100`% of their clade annotation. If only these duplication events closest to their annotated clade nodes are retained, the pattern of lower correlation of Tau evolution for duplication events disappears for the dataset simulated under the null model (*SI Appendix*, Figure 7*B*). This correction, though, comes at the high cost of discarding `r round( (n_dup_original - n_dup_remaining )/ n_dup_original * 100, 1 )`% of duplication nodes.

@@

Once we addressed the issues above with the implementation of pairwise comparisons, we were able to explore more general issues that can be a problem when making multiple pairwise comparisons between species. We found that all five of the lowest values in the mid phase distribution (Figure 3*A*) are for pairwise comparisons that include the ctenophore (comb jelly). When the nine pairwise comparisons that include the ctenophore are removed, there is no significant difference between the early phase and mid phase distributions (p = 0.14 for the early-mid comparison and p < $10^{-5}$ for the late-mid comparison) and no support for the inverse hourglass (Figure 3*B*). This highlights a well understood property of pairwise comparisons across species [@Felsenstein:1985ua; @Ackerly:1999iu]: evolutionary changes along a given branch, like those along the ctenophore branch, impact each of the multiple pairwise comparisons that includes that branch. The pairwise comparisons are therefore not independent - different pairwise comparisons are impacted by changes along some of the same branches (Figure 1*A*). This can give the impression of a general pattern across the tree that is instead specific to changes along one part of the tree. The number of comparisons impacted by each change depends on the structure of the phylogenetic tree, *i.e.* how the species are related to each other. Phylogenetic comparative methods were developed specifically to address this problem [@Felsenstein:1985ua].

@@

After parsing the trees from the Compara file with treeio, which was recently split from ggtree [@Yu:2016fo], we added Tau estimates generated by the KMRR `Rscript.R` to the tree data objects. We then pruned away tips without expression data, retaining only the trees with `r min_genes_with_expression %>% format(big.mark=",")` or more tips. We also only retained trees with one or more speciation events, as speciation events are required for calibration steps. This removes trees that have multiple genes from only one species after pruning away tips without expression data.

@@

```{r within_species_variation, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.width=6.5, fig.height=3, fig.cap="Investigation of the potential impact of within-species variance. These plots correspond to Figures 2*A* and 2*B*, but with increased expected within-species variation modeled by increasing the terminal gene phylogeny branch lengths. (*A*) Pairwise comparisons still recover higher correlation across speciation events than duplication events. (*B*) There is not an excess of larger contrasts for duplication events and the analysis does not reject the null hypothesis. Modeling a large amount of within-species variation does not lead to support for the ortholog conjecture."}
